MESOSCOPE manuscript results/figures

### 2023-08-29 manuscript/manuscript.Rmd
All-new results section completed

---
title: "AllMeso manuscript"
author: "William Kumler"
date: "`r Sys.Date()`"
output: html_document
---

```{r Run analysis}
if(!file.exists("../metadata/filled_file_metadata.csv")){
  setwd("../metadata")
  knitr::purl("metadata_control.Rmd")
  source("metadata_control.R")
  file.remove("metadata_control.R")
  setwd("../manuscript")
}
if(!file.exists("../untargeted/all_peaks.csv")){
  setwd("../untargeted")
  data_iters <- c("FK_pos", "FK_neg", "MS_pos", "MS_neg")
  knitr::purl("untargeted_control.Rmd")
  for(cruise in c("FK", "MS")){
    for(polarity in c("pos", "neg")){
      source("untargeted_control.R")
    }
  }
  file.remove("untargeted_control.R")
  map(data_iters, function(output_dir){
    comb_peaks <- read_csv(paste0(output_dir, "_output/clean_peak_data.csv")) %>%
      mutate(temp=output_dir) %>%
      separate(temp, into = c("cruise", "polarity"), sep = "_") %>%
      mutate(feature=str_replace(feature, "FT", paste0(toupper(output_dir), "_")))
  }) %>%
    bind_rows() %>%
    write_csv("all_peaks.csv")
  map(data_iters, function(output_dir){
    comb_peaks <- read_csv(paste0(output_dir, "_output/clean_feat_data.csv")) %>%
      mutate(temp=output_dir) %>%
      separate(temp, into = c("cruise", "polarity"), sep = "_") %>%
      mutate(feature=str_replace(feature, "FT", paste0(toupper(output_dir), "_")))
  }) %>%
    bind_rows() %>%
    write_csv("all_feats.csv")
  setwd("../manuscript")
}
if(!file.exists("../targeted/all_peaks.csv")){
  setwd("../targeted")
  knitr::purl("../targeted/targeted_control.Rmd")
  source("targeted_control.R")
  file.remove("targeted_control.R")
  setwd("../manuscript")
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

filled_file_metadata <- read_csv("../metadata/filled_file_metadata.csv")
clean_stans <- read_csv("../metadata/clean_stans.csv")
untarg_feats <- read_csv("../untargeted/all_feats.csv")
untarg_peaks <- read_csv("../untargeted/all_peaks.csv")
targ_peaks <- read_csv("../targeted/all_peaks.csv")
```

>>>>

---
title: "MESOSCOPE manuscript results/figures"
author: "William Kumler"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
---

```{r Run analysis, include=FALSE}
if(!file.exists("../mzMLs/neg/180205_Poo_TruePoo_Full1.mzML")){
  init_timeout <- options("timeout")$timeout
  options(timeout=3600) # Allow 1 hour for each FTP download

  download.file("ftp://www.metabolomicsworkbench.org/Studies/ST002788_HILIC_NEG.zip",
                destfile = "../mzMLs/ST002788_HILIC_NEG.zip")
  unzip("../mzMLs/ST002788_HILIC_NEG.zip", exdir = "../mzMLs/neg")
  file.remove("../mzMLs/ST002788_HILIC_NEG.zip")
  download.file("ftp://www.metabolomicsworkbench.org/Studies/ST002789_HILIC_NEG.zip",
                destfile = "../mzMLs/ST002789_HILIC_NEG.zip")
  unzip("../mzMLs/ST002789_HILIC_NEG.zip", exdir = "../mzMLs/neg")
  file.remove("../mzMLs/ST002789_HILIC_NEG.zip")
  download.file("ftp://www.metabolomicsworkbench.org/Studies/ST002788_HILIC_POS.zip",
                destfile = "../mzMLs/ST002788_HILIC_POS.zip")
  unzip("../mzMLs/ST002788_HILIC_POS.zip", exdir = "../mzMLs/pos")
  file.remove("../mzMLs/ST002788_HILIC_POS.zip")
  download.file("ftp://www.metabolomicsworkbench.org/Studies/ST002789_HILIC_POS.zip",
                destfile = "../mzMLs/ST002789_HILIC_POS.zip")
  stop("File extraction has to be done manually - file too large for R's unzip")
  file.remove("../mzMLs/ST002789_HILIC_POS.zip")
  options(timeout=init_timeout)
}

if(!file.exists("../metadata/filled_file_metadata.csv")){
  setwd("../metadata")
  knitr::purl("metadata_control.Rmd")
  source("metadata_control.R")
  file.remove("metadata_control.R")
  setwd("../manuscript")
}

if(!file.exists("../targeted/all_peaks.csv")){
  setwd("../targeted")
  knitr::purl("../targeted/targeted_control.Rmd")
  source("targeted_control.R")
  file.remove("targeted_control.R")
  setwd("../manuscript")
}

if(!file.exists("../untargeted/all_peaks.csv")){
  setwd("../untargeted")
  knitr::purl("untargeted_control.Rmd")
  data_iters <- c("FK_pos", "FK_neg", "MS_pos", "MS_neg")
  for(cruise in c("FK", "MS")){
    for(polarity in c("pos", "neg")){
      source("untargeted_control.R")
    }
  }
  file.remove("../untargeted/untargeted_control.R")
  map(data_iters, function(output_dir){
    comb_peaks <- read_csv(paste0(output_dir, "_output/clean_peak_data.csv")) %>%
      mutate(temp=output_dir) %>%
      separate(temp, into = c("cruise", "polarity"), sep = "_") %>%
      mutate(feature=str_replace(feature, "FT", paste0(toupper(output_dir), "_")))
  }) %>%
    bind_rows() %>%
    write_csv("all_peaks.csv")
  map(data_iters, function(output_dir){
    comb_peaks <- read_csv(paste0(output_dir, "_output/clean_feat_data.csv")) %>%
      mutate(temp=output_dir) %>%
      separate(temp, into = c("cruise", "polarity"), sep = "_") %>%
      mutate(feature=str_replace(feature, "FT", paste0(toupper(output_dir), "_")))
  }) %>%
    bind_rows() %>%
    write_csv("all_feats.csv")
  setwd("../manuscript")
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)

filled_file_metadata <- read_csv("../metadata/filled_file_metadata.csv") %>%
  filter(polarity=="pos") %>%
  select(-polarity)
clean_stans <- read_csv("../metadata/clean_stans.csv")
untarg_feats <- read_csv("../untargeted/all_feats.csv")
untarg_peaks <- read_csv("../untargeted/all_peaks.csv") %>%
  filter(str_detect(filename, "_Smp_"))
targ_peaks <- read_csv("../targeted/all_peaks.csv") %>%
    mutate(filename=paste0(filename, ".mzML"))

metadata_levels <- c(
    "no23_um", "po4_um", "si_um", "PC_um", "PN_um", "DON_um", "DOP_um", "oxy_um",
    "chl_ug", "phaeo_ug", "alk", "dic", "ph", "theta",
    paste0(c("pro", "syn", "euk", "het"), "_abund"), "beam_atten",
    "abs_depth", "time", "lat", "sla", "local_hour", 
    "TDN", "TDP", "temp", "pressure", "sal", "sigma",
    "ctd_chl"
  )
metadata_labels <- c(
    "Nitrate + nitrite", "Phosphate", "Silicate", "Particulate C", "Particulate N",
    "Dissolved organic N", "Dissolved organic P", "Oxygen", "Chlorophyll a",
    "Phaeophytin", "Alkalinity", "Dissolved inorganic C", "pH", "Density",
    paste(c("Pro.", "Syn.", "Picoeuk.", "Het. bact."), "abundance"),
    "Beam attenuation",
    "Depth", "Time", "Latitude", "Sea level anomaly", "Local hour", 
    "Total dissolved nitrogen", "Total dissolved phosphorus",
    "Temperature", "Pressure", "Salinity", "Sigma-theta",
    "Chlorophyll (fluor.)"
  )
metadata_twoline_labels <- c(
    "Nitrate + nitrite", "Phosphate", "Silicate", "Particulate C", "Particulate N",
    "Dissolved\norganic N", "Dissolved\norganic P", "Oxygen", "Chlorophyll a",
    "Phaeophytin", "Alkalinity", "Dissolved\ninorganic C", "pH", "Density",
    paste(c("Prochlorococcus", "Synechococcus", "Picoeukaryote", 
            "Heterotrophic bact."), "\nabundance"),
    "Beam attenuation",
    "Depth", "Time", "Latitude", "Sea level anomaly", "Local hour",
    "Total dissolved\nnitrogen", "Total dissolved\nphosphorus",
    "Temperature", "Pressure", "Salinity", "Sigma-theta", "Chlorophyll\n(fluorescence)"
  )
metadata_short_labels <- c("N+N", "PO4", "Si", "PC", "PN", "DON", "DOP", "O2", "Chl a",
                  "Phaeophytin", "Alk", "DIC", "pH", "Density", 
                  paste(c("Pro", "Syn", "Picoeuk", "Het. bact."), "abund"),
                  "Beam atten.", "Depth", "Time", "Latitude", "SLA", "Local hour",
                  "Tot. N", "Tot. P", "Temp", "dB", "Sal", "σ-θ", "CTD chl.")

splividis <- function(n, ...){
  vir <- viridis::viridis(n, ...)
  if(n%%2==0){
    vir[as.numeric(matrix(1:n, ncol=2, byrow = TRUE))]
  } else {
    index_vec <- suppressWarnings(as.numeric(matrix(1:n, ncol=2, byrow = TRUE)))
    vir[head(index_vec, -1)]
  }
}
```

### Metabolome changes significantly across the eddy dipole at DCM and 175m

We found that the metabolome differed between the cyclone and anticyclone samples, with multivariate statistics detecting the largest differences in the DCM and 175 meter samples while the surface was more homogenous (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by SLA alone according to the PERMANOVA (p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the deep samples, 12% of the variance was explained by SLA (p = 0.008) according to the PERMANOVA and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to capture largely variation between biological triplicates. At the surface, SLA trends are less evident with a larger p-value (0.087) and a lower R$^2$ (0.09) (Figure 1).

To characterize this difference, we plotted an "average" metabolite per station at each depth by taking the median value of all the detected mass features after z-score normalization (Figure 1). As expected, the average metabolite collected at the exact center of the cyclonic eddy (Station 12) was much more abundant than in any sample at this same depth in the rest of the dipole samples. Similarly, the 175 meter samples showed the largest metabolite abundances in the anticyclone relative to other stations at this depth. This method also allowed us to see the remarkably sinusoidal response of the average metabolite across the stations. This response correlated negatively with SLA at the DCM (lower SLA = higher average values, $r_{DCM}= -.716$) but positively with SLA at the surface and at depth ($r_{surface} = .211$, $r_{deep} = .830$).

  - Discussion: note that the "neither" samples all group together, implying eddy effect rather than location ("outside" the eddy is more similar to each other than to the insides of either one)
  - Discussion: it's kinda crazy that the exact center of the eddy is so wildly different at the DCM from stations that are nearby but not

```{r multivar_analysis object creation - nmds/pca/permanova}
set.seed(123)
multivar_analysis <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  filter(ms_run=="MT") %>%
  nest(data=-depth) %>%
  mutate(wide_peaks=map(data, function(metab_df){
    metab_df %>%
      select(feature, filename, norm_area) %>%
      group_by(feature) %>%
      filter(sum(norm_area==0)<n()*0.9) %>%
      mutate(norm_area=scale(norm_area)[,1]) %>%
      pivot_wider(names_from = "feature", values_from = norm_area) %>%
      column_to_rownames("filename") %>%
      data.matrix()
  })) %>%
  mutate(nmds_output=map(wide_peaks, function(metab_mat){
    env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat)]
    vegan::metaMDS(metab_mat, k = 2, distance = "manhattan", trace = 0, 
                   autotransform = FALSE, noshare = FALSE, wascores = FALSE) %>%
    vegan::MDSrotate(env_vec)
  })) %>%
  mutate(nmds_stress=map_dbl(nmds_output, pluck, "stress")) %>%
  mutate(perm_output=map(wide_peaks, function(metab_mat){
    adon_expl <- data.frame(filename=rownames(metab_mat)) %>%
      left_join(filled_file_metadata %>% select(filename, sla_class), by="filename")
    vegan::adonis2(
      metab_mat~sla_class, data = adon_expl, by = "margin", 
      permutations = 999, method = "manhattan")
  })) %>%
  mutate(perm_r2=map_dbl(perm_output, pluck, "R2", 1)) %>%
  mutate(perm_pval=map_dbl(perm_output, pluck, "Pr(>F)", 1)) %>%
  mutate(pca_output=map(wide_peaks, prcomp)) %>%
  mutate(nmds_stress=paste("s =", round(nmds_stress, digits = 3))) %>%
  mutate(perm_r2=round(perm_r2, digits = 2)) %>%
  mutate(perm_pval=round(perm_pval, digits = 4)) %>%
  mutate(perm_label=paste0("R2 = ", perm_r2, "\np = ", perm_pval))
```

```{r figure nmds and median metabolite across eddy transect}
nmds_sla_gp <- multivar_analysis %>%
  mutate(nmds_points=map(nmds_output, function(nmds_output_i){
    nmds_output_i %>% pluck("points") %>% as.data.frame()
  })) %>%
  select(depth, nmds_points) %>%
  unnest_longer(nmds_points, indices_to = "filename") %>%
  unnest_wider(nmds_points) %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_corr)) %>%
  mutate(MDS2=ifelse(str_detect(depth,"Deep"), MDS2*-1, MDS2)) %>%
  ggplot() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, fill=sla_corr), color="black", size=4, pch=21) +
  geom_label(x=-Inf, y=-Inf, aes(label=nmds_stress), hjust=0, vjust=0, 
             data = multivar_analysis, label.r = unit(0, "in")) +
  geom_label(x=-Inf, y=Inf, aes(label=perm_label), hjust=0, vjust=1,
             data = multivar_analysis, label.r = unit(0, "in"), lineheight = .7) +
  facet_wrap(~depth, nrow=1) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill"),
  ) +
  labs(fill="Sea level\nanomaly\n(cm)\n") +
  theme_bw() +
  coord_fixed()

med_metab_df <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(filename, ms_run, station, depth, sla_corr),
            by="filename") %>%
  filter(ms_run=="MT") %>%
  group_by(depth, feature) %>%
  filter(sum(norm_area==0)<n()*0.9) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  ungroup() %>%
  group_by(filename, depth, station, sla_corr) %>%
  summarize(avg_amt=mean(norm_area)) %>%
  ungroup()
# Calculations for correlation between grand median metabolite and SLA
# med_metab_df %>%
#   group_by(station, depth, sla_corr) %>%
#   summarise(grand_avg_amt=mean(avg_amt)) %>%
#   group_by(depth) %>%
#   summarise(sim=cor(sla_corr, grand_avg_amt))
med_metab_sla_gp <- med_metab_df %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_boxplot(aes(x=station, y=avg_amt, color=sla_corr), linewidth=0.8) +
  facet_wrap(~depth, nrow=1) +
  scale_color_gradient2(low = scales::muted("blue"), high = scales::muted("red"),
                        mid="grey90", aesthetics = c("color", "fill"), guide = NULL) +
  theme_bw() +
  theme(strip.background = element_blank(), strip.text = element_blank()) +
  labs(x="Station", y="Median metabolite z-score")

library(egg)
nolegend_plot <- ggarrange(nmds_sla_gp+theme(legend.position = "none"), 
                           med_metab_sla_gp, ncol = 1, heights = c(0.7, 0.3), 
                           draw = FALSE)
library(cowplot)
nmds_and_med_metab <- plot_grid(nolegend_plot, get_legend(nmds_sla_gp), 
                                nrow=1, rel_widths = c(0.9, 0.1))
ggsave("nmds_and_med_metab.png", plot = nmds_and_med_metab, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300,
       bg="white")
```

![](figures/nmds_and_med_metab.png)

*Figure 1: Plots of the metabolome across the eddy dipole, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. NMDS stress values (s) have been reported in the bottom left corner, while PERMANOVA R$^2$ and p-values are reported in the top left. SLA trends are visible in the DCM and deep samples, with dark blue circles consistently discriminating from the dark red circles along the first multidimensional axis. The bottom row of plots show the direction and magnitude of this effect by plotting the grand mean of the normalized metabolite peak areas in the stations taken across the eddy dipole.*

```{r supp figure/table prcomp vs SLA}
pcdata <- multivar_analysis %>%
  select(depth, pca_output) %>%
  mutate(pca_x=map(pca_output, function(pca_output_i){
    pca_output_i$x[,1:3] %>% as.data.frame()
  })) %>%
  select(depth, pca_x) %>%
  unnest_longer(pca_x, indices_to = "filename") %>%
  unnest_wider(pca_x) %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_corr)) %>%
  pivot_longer(starts_with("PC"))
pcvars <- multivar_analysis %>%
  select(depth, pca_output) %>%
  mutate(perc_expl=map(pca_output, function(pca_output_i){
    var_expl <- pca_output_i$sdev^2
    data.frame(perc_expl=var_expl[1:3]/sum(var_expl), name=paste0("PC", 1:3))
  })) %>%
  unnest(perc_expl) %>%
  mutate(perc_label=paste0(round(perc_expl, digits = 3)*100, "%"))

# Correlation between PCs and SLA at each depth
# pcdata %>%
#   nest(data=-c(depth, name)) %>%
#   mutate(cor_output=map_dbl(data, ~abs(cor(.x$value,.x$sla_corr))))

# Linear model significance testing and parameter report
# pcdata %>%
#   nest(data=-c(depth, name)) %>%
#   mutate(lm_output=map(data, ~broom::tidy(lm(.x$value~.x$sla_corr)))) %>%
#   unnest(lm_output) %>%
#   filter(term==".x$sla_corr") %>%
#   arrange(name, depth) %>%
#   select(depth, name, p.value) %>%
#   split(.$name)

lm_gp <- ggplot(pcdata, aes(x=sla_corr, y=value)) +
  geom_smooth(method="lm", formula=y~x, color="black") +
  geom_point(aes(fill=sla_corr), pch=21, color="black") +
  geom_label(data = pcvars, aes(label=perc_label),
             hjust=1, vjust=1, x=Inf, y=Inf, label.r = unit(0, "in")) +
  facet_grid(depth~name, switch = "y") +
  scale_y_continuous(position = "right") +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill"), guide = NULL
  ) +
  labs(x="Corrected SLA, in cm", y="Principal component value",
       color="Corrected\nSLA (cm)") +
  theme_bw()
deep_pc1_barplot_gp <- pcdata %>%
  left_join(filled_file_metadata %>% distinct(filename, station, tripl)) %>%
  mutate(station=factor(station, levels=as.character(4:14),
                        labels=paste0("St.", 4:14))) %>%
  arrange(station) %>%
  filter(name=="PC1" & str_detect(depth, "Deep")) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=tripl, y=value)) +
  facet_wrap(~station, nrow = 1, scales="free_x") +
  scale_y_continuous(position = "right") +
  labs(x="Triplicate", y="PC1", title = paste(
    "    Variance among deep (175 meter) samples in PC1"
  ))
pca_sla_cor_gp <- cowplot::plot_grid(lm_gp, deep_pc1_barplot_gp, 
                                     rel_heights = c(0.6, 0.4), 
                                     ncol = 1, labels = "AUTO")
ggsave("pca_sla_cor_gp.png", pca_sla_cor_gp, device = "png", path = "supplement",
       width = 6.5, height = 6.5, units = "in", dpi = 300)

```

![](supplement/pca_sla_cor_gp.png)

*Supplemental figure: Results of a principal component analysis performed separately on each depth of the eddy transect metabolome. A) Regressions of SLA against various principal components, broken down by depth. % variance explained by each PC is noted in the upper right corner of the plot. Colors have been added corresponding to the SLA value for clarity but are redundant with the x-axis values. Lines of best fit and the associated standard error have been added behind the points in black and translucent grey, respectively. Strong associations exist between PC1 and SLA in the DCM samples and PC2 and SLA in the deep samples. B) Principal component 1 values for the 175 meter samples at each station showing large variance between triplicates rather than any external factor.*

A large part of this metabolic response to SLA is due to the increased biomass in the cyclone DCM and the parallel increase in the anticyclone deep samples. We used a subset of the prior dataset, only those metabolites that we were able to quantify, and measured their in-situ concentrations in the particulate matter (Figure 2).

```{r figure stacked barplot MT eddy transect w PC and SLA}
avg_targ_concs <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup()
base_targ_gp <- avg_targ_concs %>%
  group_by(compound_name, untripl) %>%
  summarise(nM=mean(nM), mean_cmpd_nm=unique(mean_cmpd_nm)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Dimethylsulfoniopropionate", "DMSP", compound_name)) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 11), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  left_join(filled_file_metadata %>% distinct(untripl, depth, station)) %>%
  group_by(depth, station, compound_name) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  summarise(mean_nM=mean(nM)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=station, y=mean_nM, fill=compound_name), width = 0.9) +
  scale_fill_manual(values = c(hcl.colors(12)[c(1,3,5,7,9,11,2,4,6,8,10)], "grey50")) +
  scale_y_continuous(expand = c(0.005, 0.1)) +
  labs(x="Station", y="Mean nM metabolite concentration", fill=NULL) +
  facet_grid(~depth, scales="free", space = "free") +
  theme_bw() +
  theme(legend.position = "bottom")

sla_loess_points <- filled_file_metadata %>%
  filter(samp_type=="Smp") %>%
  filter(ms_run=="MT") %>%
  distinct(station, sla_corr) %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  group_by(station) %>%
  summarise(sla_corr=mean(sla_corr)) %>%
  mutate(station=as.numeric(station)) %>%
  loess(formula = sla_corr~station) %>%
  predict(., newdata=data.frame(station=seq(1, 11, 0.01))) %>%
  data.frame(station=seq(1, 11, 0.01), sla_corr=.)
arrow_df <- data.frame(
  cyc_type=c("Anticyclone\ncenter", "Cyclone\ncenter"),
  x=c("6", "12"), xend=c("6", "12"),
  y=c(5, 4), yend=c(7.5, 6.5),
  color=c(scales::muted("red"), scales::muted("blue")),
  depth=factor("Deep", levels=c("Surface", "DCM", "Deep"), 
               labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)")),
  lab_x=c("5", "11")
)
targ_gp_w_sla <- base_targ_gp +
  ggnewscale::new_scale_fill() +
  geom_tile(aes(x=station, y=-0.5, fill=sla_corr),
            color=NA, width=0.11, height=0.5, data = sla_loess_points) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("fill"), guide = "none"
  ) +
  geom_segment(aes(x=x, y=y, xend=xend, yend=yend, color=color), data=arrow_df,
               arrow=arrow(ends = "first", length = unit(0.1, "npc"), type = "closed"),
               linewidth=0, linejoin = "bevel") +
  geom_segment(aes(x=x, y=y+0.5, xend=xend, yend=yend, color=color), data=arrow_df, linewidth=2) +
  geom_text(aes(x=lab_x, y=yend+0.5, color=color, label=cyc_type), 
            data=arrow_df, vjust=0, hjust=0, lineheight=0.8) +
  scale_color_identity()

PC_data <- filled_file_metadata %>%
  filter(samp_type=="Smp") %>%
  filter(ms_run=="MT") %>%
  distinct(station, PC_um, depth) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15m)", "DCM  (110-120m)", "Deep  (175m)"))) %>%
  mutate(station=factor(station, levels=as.character(4:14)))
targ_gp_w_sla_PC <- targ_gp_w_sla +
  geom_point(aes(x=station, y=PC_um*5), data=PC_data, pch=21, fill="white", color="black") +
  scale_y_continuous(sec.axis = sec_axis(~./5, name = "Particulate carbon (o, µM)"),
                     expand = c(0.005, 0.1))

ggsave("targ_gp_w_sla_PC.png", plot = targ_gp_w_sla_PC, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300)
```

![](figures/targ_gp_w_sla_PC.png)

*Figure 2: Concentrations of known compounds across the eddy dipole, broken down by depth. Stacked bars correspond to known compounds named in the legend underneath, with compounds not in the top 11 by absolute concentration grouped into the "Other" category. DMSP = dimethylsulfoniopropionate. Sea level anomaly (SLA) is shown by the red-blue color scale underneath the barplot with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone). The eddy centers (stations 6 and 12) have been labeled with arrows of the same color scheme in the final panel. Particulate carbon concentrations in µmol/L are shown as white dots and use the y-axis scale on the right side (200x the metabolite scale).*

```{r avg targ conc values and numbers for text, eval=FALSE}
avg_targ_concs %>%
  group_by(filename) %>%
  summarise(nM=sum(nM)) %>%
  left_join(filled_file_metadata %>% distinct(filename, untripl, PC_um, depth)) %>%
  group_by(untripl, PC_um, depth) %>%
  summarise(nM=mean(nM), .groups = "drop") %>%
  group_by(depth) %>%
  summarise(PC_vs_nM=broom::tidy(cor.test(PC_um, nM)))
  # summarise(PC_vs_nM=broom::tidy(cor.test(PC_um, nM, method = "spearman")))
  # ggplot(aes(x=PC_um, y=nM, color=depth)) + geom_point() + geom_smooth(method="lm")
avg_targ_concs %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  group_by(depth, compound_name) %>%
  summarise(nM=mean(nM)) %>%
  arrange(compound_name) %>%
  pivot_wider(names_from = depth, values_from = nM) %>%
  write_csv("supplement/avg_targ_conc_by_depth.csv")
avg_targ_concs %>%
  mutate(depth=factor(depth, levels=rev(c("Surface", "DCM", "Deep")))) %>%
  group_by(compound_name, depth) %>%
  summarise(nM=mean(nM)) %>%
  mutate(nM=log10(nM)) %>%
  ggplot(aes(x=nM, y=depth, group=compound_name)) +
  geom_path()
avg_targ_concs %>%
  mutate(depth=factor(depth, levels=rev(c("Surface", "DCM", "Deep")))) %>%
  group_by(compound_name, depth) %>%
  summarise(nM=mean(nM)) %>%
  mutate(nM=rank(nM)) %>%
  ungroup() %>%
  pivot_wider(names_from = depth, values_from = nM) %>%
  mutate(depth_class=paste(Surface, DCM, Deep)) %>%
  count(depth_class)
```

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), representing an average concentration of 1.38 nM. At the surface, however, hydroxyisoleucine was the most concentrated metabolite at 1.53 nM due to its monotonic decrease with depth as opposed to a majority of the osmolytes (37/53) which reached maximum concentrations at the DCM instead (Figure 2).

Bulk particulate carbon (PC) measurements correlated well with overall metabolite concentrations ($r= .554$, p-value < 0.001) but this was largely driven by the lower biomass and metabolite abundance at depth. When analyzed at each depth separately, the DCM ($r_{DCM}= .790$, p-value = 0.003) and 175 meter samples ($r_{Deep}= .554$, p-value = 0.077) had weaker but still good correlations, but no such association was detected at the surface ($r_{Surface}= -.125$, p-value = 0.714).

```{r supp figure fraction of PC in metabolites}
pc_comp_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))

# Significance testing between depths and SLA classes for % PC in metab
# pc_comp_df %>% aov(formula=frac_in_metab~depth) %>% TukeyHSD()
# pc_comp_df %>% aov(formula=frac_in_metab~sla_class) %>% TukeyHSD()

# Significance testing for differences in % in metabs as a function of SLA by depth
# pc_comp_df %>%
#   nest(data=-depth) %>%
#   mutate(aov_out=map(data, ~broom::tidy(aov(.x$frac_in_metab~.x$sla_class)))) %>%
#   mutate(kw_out=map(data, ~broom::tidy(kruskal.test(.x$frac_in_metab~.x$sla_class)))) %>%
#   pivot_longer(c(aov_out, kw_out)) %>%
#   unnest(value) %>%
#   drop_na(p.value) %>%
#   select(depth, name, p.value) %>%
#   group_by(name) %>%
#   mutate(p.value=p.adjust(p.value, method="fdr")) %>%
#   ungroup() %>%
#   arrange(name, depth)
# pc_comp_df %>%
#   ggplot(aes(x=sla_class, y=frac_in_metab)) +
#   geom_boxplot() +
#   facet_wrap(~depth, nrow=1)

# Permutation testing for effect size test on Deep SLA differences
# set.seed(123)
# map_dbl(seq(1, 2, 0.1), function(n_mult){
#   map_dbl(1:100, function(h){
#     pc_comp_df %>%
#       filter(depth=="Deep") %>%
#       group_by(sla_class) %>%
#       summarize(mean=mean(frac_in_metab), sd=sd(frac_in_metab), n=n()) %>%
#       mutate(sim_vals=map(1:3, function(i){rnorm(.$n[i]*n_mult, .$mean[i], .$sd[i])})) %>%
#       unnest_longer(sim_vals) %>%
#       aov(formula=sim_vals~sla_class) %>%
#       broom::tidy() %>%
#       filter(term=="sla_class") %>%
#       pull(p.value)
#   }) %>% mean()
# })

pc_frac_metab_gp <- ggplot(pc_comp_df, aes(x=station, y=frac_in_metab, color=sla_corr)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(width = 0.9, linewidth=0.8) +
  geom_point(color="black", size=2) +
  scale_y_continuous(labels = scales::label_percent()) +
  facet_grid(~depth, scales="free") +
  scale_color_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill")
  ) +
  guides(color=guide_colorbar(direction = "horizontal", title.position = "top", 
                              title.hjust = 0.5)) +
  labs(color="Sea level\nanomaly (cm)", x="Station",
       y="Fraction of PC represented by known metabolites") +
  theme_bw() +
  theme(legend.position = c(0.85, 0.8), legend.background = element_rect(color="grey90"))
ggsave("pc_frac_metab_gp.png", pc_frac_metab_gp, device = "png", path = "supplement",
       width = 6.5, height = 3.5, units = "in", dpi = 300)
```

![](supplement/pc_frac_metab_gp.png)

*Supplemental figure: Fraction of particulate carbon represented by known metabolites. Boxplots set behind points are colored according to sea level anomaly (SLA) with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone).*

This correlation between PC and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. Known metabolites contributed between approximately 1.5% and 5% of the carbon biomass to the total PC load, but this variance was not clearly due to either depth or SLA. The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.3-0.4 percentage points lower at depth, p-value=0.018) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). However, this lack of difference observed may have been due to the small sample sizes (18 anticyclonic samples, 9 cyclonic, and 6 neither). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for the 175 meter samples if 1.5x as many samples had been collected with mean and variances remaining similar.
